A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2021-04-20 submission/manuscript.rmd
new version of manuscript

---
title: "A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases"
output:
    bookdown::pdf_document2: default
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left, pagewise]{lineno}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jamia.csl
---

\newpage
# Abstract (247 words)
\singlespacing 
**Aim:** One of the aims of the Observation Health Data Sciences
and Informatics (OHDSI) initiative is population-level treatment effect
estimation in large observational databases. Since treatment effects are
well-known to vary across groups of patients with different baseline risk, we
aimed to extend the OHDSI methods library with a framework for risk-based
assessment of treatment effect heterogeneity.

@@

The Observational Health Data Sciences and Informatics (OHDSI) collaborative
(!!REF) has established an international network of data partners and
researchers that aim to bring out the value of health data through large-scale
analytics by mapping all available databases to the Observational Medical
Outcomes Partnership (OMOP) Common Data Model (CDM) [@Overhage2012]. The
common data structure enables analyses at a very large scale. For example, in a
recent study [@Suchard2019], a large set of first-line treatments for
hypertension was compared with respect to 55 outcomes in a network of databases,
including 4.9 million patients from around the world.

@@

# Methods
The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the database within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation and evaluation of the results. A simple overview of the procedure
can be seen in Figure XXXX.

@@

# Results
As a proof of concept, we focus on the comparison of angiotensin converting
enzyme (ACE) inhibitors to beta blockers. ACE inhibitors are among the most
common treatment classes for hypertension, with a well-established
effectiveness. Beta blockers, even though initially widely used for the
treatment of hypertension, more recent trials and meta-analyses have cast doubt
on their relative effectiveness (!!REF). As a result, newer US guidelines do
not consider them for initial treatment for hypertension while in the EU
guidelines combination with other antihypertensive treatments is recommended
[@Whelton2018; @Williams2018]. However, another meta-analysis suggested that
the efficacy profile of beta blockers is similar to other major treatment
classes in younger hypertensive patients and, thus, countries like Canada still
include them as a first-line candidate for the treatment [@Khan2006;
@Rabi2020].

@@

For each outcome and in each risk stratum there were adequate numbers of
patients (Table XXXX). The discriminative ability of the prediction models was
moderate in the matched development subset (c-index 0.76 for acute MI ; 0.79 for
hospitalization with heart failure;  0.74 for stroke; ), in the general
population (c-index 0.74 for acute MI; 0.77 for hospitalization with heart
failure; 0.73 for stroke), in the treatment cohort (c-index for acute MI it was
0.71, for hospitalization with heart failure was 0.76 and for stroke it was
0.72) and in the comparator cohort (c-index for acute MI it was 0.79 for
hospitalization with heart failure was 0.79 and for stroke it was 0.75).

Relative treatment effects of ACE-inhibitors vs beta blockers increased (hazard
ratios decreased) with increasing acute MI risk, resulting in more pronounced
increases of absolute treatment effects (ARD) with increasing acute MI risk
(Figure XXXX). Patients in the low risk quarter did not receive absolute
treatment benefit (ARD -0.03%) while absolute risk was 0.54% lower (95%
confidence interval 0.36%—0.71%) for patients in the high-risk quarter. In
contrast, the absolute and relative effects of ACE-inhibitors on safety outcomes
(e.g. cough and angioedema) are approximately constant or even slightly
decreasing with increasing acute MI risk (Figure XXXX and XXXX). Similar results
were observed in the other two databases (see supplementary material).

@@

Our contribution is a translation of the PATH statement principles to the OHDSI
methods library. Our methods encourage open science as it requires clear
definition of the research questions translated into clear and reproducible
cohort definitions that can easily be shared among researchers. Our R-package
provides a standardized stepwise procedure for the assessment of HTE. This
enables source code to be easily shared and evaluated. The results of these
analyses can be reproduced in a straightforward manner. Researchers with access
to different databases mapped to OMOP-CDM can also very easily extend their
overall analyses with risk-based assessment of treatment effect heterogeneity.
This enables collaboration among multiple sites with access to different patient
populations. We propose that the framework is implemented any time treatment
effect estimation in high-dimensional observational data is undertaken.

@@

\newpage
# References
\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent
<div id="refs"></div>
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\noindent

>>>>

---
title: "A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases"
author: 
  - Alexandros Rekkas, MSc${^1}$ 
  - David van Klaveren, PhD$^{2,3}$, 
  - Patrick B. Ryan, PhD$^4$
  - Ewout W. Steyerberg, PhD$^{3,5}$
  - David M. Kent, PhD$^2$
  - Peter R. Rijnbeek, PhD${^1}$
output:
    bookdown::pdf_document2: default
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left, pagewise]{lineno}
  - \usepackage{caption}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jamia.csl
---

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(here)
library(kableExtra)
d <- function(x, decimals = 2) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}
knit_hooks$set(
  inline = function(x) {
    prettyNum(
      x,
      big.mark = ",",
      decimal.mark = ".",
      preserve.width = "individual"
    )
  }
)
```
\vspace{20mm}

$^1$ Department of Medical Informatics, Erasmus University Medical Center,
Rotterdam, Netherlands

$^2$ Predictive Analytics and Comparative Effectiveness (PACE) Center, Institute
for Clinical Research and Health Policy Studies (ICRHPS), Tufts Medical Center,
Boston, MA, USA

$^3$ Department of Public Health, Erasmus University Medical Center, Rotterdam,
Netherlands

$^4$ Janssen Research and Development, 125 Trenton Harbourton Rd,
Titusville,NJ 08560, USA

$^5$ Department of Biomedical Data Sciences, Leiden University Medical Center,
Leiden, The Netherlands

\vspace{30mm}
**Funding**: This work has been performed in the European Health Data and
Evidence Network (EHDEN) project. This project has received funding from the
Innovative Medicines Initiative 2 Joint Undertaking (JU) under grant agreement
No 806968. The JU receives support from the European Union’s Horizon 2020
research and innovation programme and EFPIA.

\newpage
\newpage
# Abstract (247 words)
\singlespacing 
**Aim:** One of the aims of the Observation Health Data Sciences
and Informatics (OHDSI) initiative is population-level treatment effect
estimation in large observational databases. Since treatment effects are
well-known to vary across groups of patients with different baseline risk, we
aimed to extend the OHDSI methods library with a framework for risk-based
assessment of treatment effect heterogeneity.

@@

The Observational Health Data Sciences and Informatics (OHDSI) collaborative
[@hripcsak2015observational] has established an international network of data
partners and researchers that aim to bring out the value of health data through
large-scale analytics by mapping all available databases to the Observational
Medical Outcomes Partnership (OMOP) Common Data Model (CDM) [@Overhage2012].
The common data structure enables analyses at a very large scale. For example,
in a recent study [@Suchard2019], a large set of first-line treatments for
hypertension was compared with respect to 55 outcomes in a network of databases,
including 4.9 million patients from around the world.

@@

# Methods
The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the database within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation and evaluation of the results. A simple overview of the procedure
can be seen in Figure \@ref(fig:figure1).

```{r figure1, echo=FALSE, fig.cap="Caption for figure 1", out.width = '100%'}
# knitr::include_graphics(here("figures/Figure_1.tif"))
grid::grid.raster(tiff::readTIFF(here("figures/Figure_1.tif")))
```

@@

# Results
As a proof of concept, we focus on the comparison of angiotensin converting
enzyme (ACE) inhibitors to beta blockers. ACE inhibitors are among the most
common treatment classes for hypertension, with a well-established
effectiveness. Beta blockers, even though initially widely used for the
treatment of hypertension, more recent trials and meta-analyses have cast doubt
on their relative effectiveness [@wiysonge2017beta]. As a result, newer US guidelines do
not consider them for initial treatment for hypertension while in the EU
guidelines combination with other antihypertensive treatments is recommended
[@Whelton2018; @Williams2018]. However, another meta-analysis suggested that
the efficacy profile of beta blockers is similar to other major treatment
classes in younger hypertensive patients and, thus, countries like Canada still
include them as a first-line candidate for the treatment [@Khan2006;
@Rabi2020].

@@

For each outcome and in each risk stratum there were adequate numbers of
patients (Table \@ref(tab:table)). The discriminative ability of the prediction
models was moderate in the matched development subset (c-index 0.76 for acute MI
; 0.79 for hospitalization with heart failure; 0.74 for stroke; ), in the
general population (c-index 0.74 for acute MI; 0.77 for hospitalization with
heart failure; 0.73 for stroke), in the treatment cohort (c-index for acute MI
it was 0.71, for hospitalization with heart failure was 0.76 and for stroke it
was 0.72) and in the comparator cohort (c-index for acute MI it was 0.79 for
hospitalization with heart failure was 0.79 and for stroke it was 0.75).
\captionsetup{width=\textwidth}
```{r table, echo=FALSE, results='asis', booktabs=TRUE}
table <- tibble(
  outcome = c(
    rep("Acute myocardial infarction", 4),
    rep("Heart failure (hosp)", 4),
    rep("Stroke (ischemic or hemorrhagic)", 4)
  ),
  riskQuarter = rep(1:4, 3),
  patientsAce = c(
    161099, 204882, 214413, 204167,
    146259, 188006, 218052, 230226,
    146069, 187524, 217070, 226128
  ),
  personYearsAce = c(
    276171, 372197, 393583, 351727,
    249809, 341014, 399394, 400330,
    294484, 340234, 397830, 393861
  ),
  eventsAce = c(
    203, 534, 117, 2095,
    228, 457, 826, 2012,
    299, 554, 947, 1718
  ),
  patientsBeta = c(
    133977, 90193, 80662, 90908,
    126387, 84280, 83421, 98380,
    126264, 84000, 83038, 97628
  ),
  personYearsBeta = c(
    220633, 169231, 150035, 154419,
    206706, 158425, 155222, 169139,
    206453, 157913, 154587, 167810
  ),
  eventsBeta = c(
    135, 321, 535, 1520,
    378, 340, 570, 1773,
    320, 351, 521, 1077
  )
)

knitr::kable(
  table,
  format = "latex",
  longtable = TRUE,
  booktabs = TRUE,
  col.names = c(
    "Outcome", "Risk quarter",
    "Persons", "Person years", "Events",
    "Persons", "Person years", "Events"
  ),
   format.args = list(
    big.mar = ","
  ),
  caption = "Number of patients, person years and events within quarters of predicted risk for hospitalization with heart failure for the 3 main outcomes of the study (acute myocardial infarction, hospitalization with heart failure and ischemic or hemorrhagic stroke)."
) %>%
  kableExtra::add_header_above(
    c(
      "", "", 
      "ACE inhibitors" = 3,
      "Beta blockers"  = 3
    )
  )
```

Relative treatment effects of ACE-inhibitors vs beta blockers increased (hazard
ratios decreased) with increasing acute MI risk, resulting in more pronounced
increases of absolute treatment effects (ARD) with increasing acute MI risk
(Figure \@ref(fig:figure2)).

```{r figure2, echo=FALSE, fig.cap="Overview of heterogeneity of ACE-inhibitors treatment within strata of predicted risk of acute MI. The top panel contains the observed acute rates of ACE-inhibitors and beta blockers within each quarter of predicted acute MI risk. These are derived using the KaplanMeier estimates at 730 days after inclusion. The middle panel, contains the hazard ratios of comparing ACE-inhibitors to beta blockers with regard to acute MI. These are estimated using Cox proportional hazards regression within quarters of predicted acute MI risk. The bottom panel contains absolute risk reduction for ACE-inhibitors compared to beta blockers. These are derived as the difference in Kaplan-Meier estimates at 730 after inclusion. Hazard ratios in the middle panel show a decreasing trend with increasing acute MI risk. Given the rather good discrimination of the prediction model (AUC=0.74), this results in an increasing trend for absolute benefit in favor of ACE-inhibitors with increasing risk.", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/Figure_2.tif")))
```
Patients in the low risk quarter did not receive absolute treatment benefit (ARD
-0.03%) while absolute risk was 0.54% lower (95% confidence
interval 0.36%—0.71%) for patients in the high-risk quarter. In contrast, the
absolute and relative effects of ACE-inhibitors on safety outcomes (e.g. cough
and angioedema) are approximately constant or even slightly decreasing with
increasing acute MI risk (Figure \@ref(fig:figure3) and \@ref(fig:figure4)).
Similar results were observed in the other two databases (see supplementary
material).

```{r figure3, echo=FALSE, fig.cap="Hazard ratios (relative treatment effects) for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI). The four risk quarters (Q1-Q4) are defined using the internally developed model for acute MI", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/Figure_3.tif")))
```
```{r figure4, echo=FALSE, fig.cap="Absolute risk reduction for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI). The four risk quarters (Q1-Q4) are defined using the internally developed model for acute MI", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/Figure_4.tif")))
```

@@

Our contribution is a translation of the PATH statement principles to the OHDSI
methods library. Our method encourages open science as it requires clear
definition of the research questions translated into clear and reproducible
cohort definitions that can easily be shared among researchers. Our R-package
provides a standardized stepwise procedure for the assessment of HTE. This
enables source code to be easily shared and evaluated, allowing for reproducible
research . Researchers with access to different databases mapped to OMOP-CDM can
also very easily extend their overall analyses with risk-based assessment of
treatment effect heterogeneity. This enables collaboration among multiple sites
with access to different patient populations. We propose that the framework is
implemented any time treatment effect estimation in high-dimensional
observational data is undertaken.

@@

\newpage
# References
\nolinenumbers
\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent
<div id="refs"></div>
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\noindent
